---
title: "Patient Survival Prediction Project"
author: "Adam Crawford, Adam Baer, Austin Funcheon, Viraj Rane"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Data

```{r}
df <- read.csv("dataset.csv")
str(df)
summary(df)

library(ggplot2)
ggplot(df, aes(factor(hospital_death))) +
  geom_bar()
```

# Data Preparation

Import data	                                                          Crawford
Store data into a dataframe	                                          Crawford
Perform high level review of data	                                    Crawford
Run prelim statistics on data	                                        Crawford
Review data for missing data	                                        Baer
Identify High integrity predictors                                    Baer
Remove unhelpful predictors (Such as, perhaps, ID)                    Baer
Transform data for usability	                                        Rane
Visualize data	                                                      Baer


Identify if any continuous data is non-normal	                        Funcheon
```{r}
#install.packages("dplyr")
library("dplyr")
library("ggpubr")
#is it a character?
ischar <- sapply(df, is.character)
ischar
df1 <- df
#visualize data
```

```{r}
#ggqqplot(df$h1_resprate_max)
df1 <- df
df1$h1_resprate_max <- log(df$h1_resprate_max)
#ggqqplot(df1$h1_resprate_max)
```



Identify if any discrete data is heavily unbalanced                   Rane
Consider data normalization	                                          Rane
Consider balancing data	                                              Funcheon
Consider transformation of some fields, such as log transformation	  Funcheon
Cleanse data                                                        	Funcheon
Format datafields into data levels (if needed)                    	  Funcheon
Run statistical analysis of predictors  .                           	Rane
Prune unhelpful predictors to drive to parsimonous model.	            Rane
Evaluate for collinearity	                                            Baer
Reduce model for predictors that have high colinearity	              Baer
Re-run statistical analysis on reduced model	                        Crawford
Break data up into train test and validate data set	                  Crawford
Run model type #1	                                                    Crawford
Evaluate results of model type #1	                                    Crawford
Run model type #2	                                                    Crawford
Evaluate results of model type #2	                                    Crawford
Run model type #3	                                                    Rane
Evaluate results of model type #3	                                    Rane
Run model type #4	                                                    Funcheon
Evaluate results of model type #4	                                    Funcheon
Tune parameters of favored model                                    	Rane
Validate favored model on Test group                                	Rane

