---
title: "Patient Survival Prediction Project"
author: "Adam Crawford, Adam Baer, Austin Funcheon, Viraj Rane"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Data

```{r}
df <- read.csv("dataset.csv")

#str(df)
#summary(df)

library("dplyr")
library("ggpubr")
library(tree)
library(caret)
library(randomForest)
library(lattice)
library(dplyr)
library(lubridate)
library(doParallel)
library(smotefamily)
library(ROSE)
cl <- makePSOCKcluster(detectCores() -1)
registerDoParallel(cl)

library(ggplot2)
#ggplot(df, aes(factor(hospital_death))) +
#  geom_bar()
```

# Data Preparation


```{r}
# get rid of identifiers and random X column
#temp removal of apache variables.
df$encounter_id <- df$patient_id <- df$hospital_id <- df$X <- df$apache_4a_hospital_death_prob <- df$apache_4a_icu_death_prob <- df$apache_3j_diagnosis <- df$apache_3j_diagnosis <- NULL

# change columns to factors
factors <- c("elective_surgery","ethnicity","gender","icu_admit_source",
             "icu_stay_type","icu_type","apache_post_operative","arf_apache",
             "gcs_eyes_apache","gcs_motor_apache","gcs_unable_apache","gcs_verbal_apache",
             "intubated_apache","ventilated_apache","aids","cirrhosis",
             "diabetes_mellitus","hepatic_failure","immunosuppression","leukemia",
             "lymphoma","solid_tumor_with_metastasis", "hospital_death")

#factors <- c("elective_surgery","ethnicity","gender","icu_admit_source",
#             "icu_stay_type","icu_type","apache_post_operative","arf_apache",
#             "gcs_eyes_apache","gcs_motor_apache","gcs_unable_apache","gcs_verbal_apache",
#             "intubated_apache","ventilated_apache","aids","cirrhosis",
#             "diabetes_mellitus","hepatic_failure","immunosuppression","leukemia",
#             "lymphoma","solid_tumor_with_metastasis","apache_3j_bodysystem",
#             "hospital_death")
df[factors] <- lapply(df[factors], factor)

#table(df$hospital_death)

df0 <- df # keeping an origin copy of df. I suspect we don't want to blanket kill all the omits if we later find those factors are irrelevant. our dataset cuts in half ~Austin



dfBal <- ovun.sample(hospital_death~., data = df, method = "over", N = 114792)$data

df0Bal <- dfBal

#summary(dfBal)

#table(df$hospital_death)
#table(dfBal$hospital_death)



```

```{r}
df <- na.omit(df)
dfBal <- na.omit(dfBal)
```



```{r}
# split data into training data and testing data
#these variables are only used for first pass. Use trim_train_data and trim_test_data instead after bottom 30 variables reduced.
set.seed(123)
train <- sample(1:nrow(dfBal), nrow(dfBal)*0.7)
train_data <- dfBal[train,]
test_data <- dfBal[-train,]
```

```{r}
library(tree)

#this doesn't work
first_tree <- tree(as.factor(hospital_death) ~., data = train_data)
summary(first_tree)
plot(first_tree)
text(first_tree, cex = 0.75, col = 'red')
```

```{r}
#Austin Fit a random forest for variable selection, run time ~7.5 minutes
library(randomForest)
#set.seed(123)
start_t <- Sys.time()
cat("",cat(" Variable selection Training started at:",format(start_t, "%a %b %d %X %Y")))

#run a prelim rf for variable summary. 
rf_eval <- randomForest(hospital_death~., data = train_data,
                           mtry = 5, importance = TRUE)

rf_eval
varImpPlot(rf_eval) #plot relevance plot

finish_t <- Sys.time()
cat("",cat("Variable selection Training finished at:",format(finish_t, "%a %b %d %X %Y")))

cat("Variable selection The training process finished in",difftime(finish_t,start_t,units="mins"), "minutes")


```

```{r} 
#Austin Identify the lowest impact variables from the RF run
rfImp <- importance(rf_eval)
#rfImp

rf_Imp_Sort <- as.data.frame.matrix(rfImp)
#adds a column for accuracy*gini for overall variable impact
rf_Imp_Sort$cross <- rf_Imp_Sort$MeanDecreaseAccuracy*rf_Imp_Sort$MeanDecreaseGini
#rf_Imp_Sort
#adds an overall rank of impact of variable, with bigger being biggest ranked impact, with 1 being lowest impact
rf_Imp_Sort$rank <- rank(rf_Imp_Sort$cross)
#rf_Imp_Sort

#variables ranked by cross order
rf_Imp_SO <- rf_Imp_Sort[order(-rf_Imp_Sort$cross),]
#rf_Imp_SO
#variable list sorted by importance. Accuracy * Gini
```

```{r} 
#Check accuracy score against test data.
rf_yhat <- predict(rf_eval, newdata = test_data)
#accuracy score of mtry 5 rf used for variable selection
rfscore <- postResample(rf_yhat, test_data$hospital_death)
print(rfscore)
```
 Accuracy     Kappa 
0.9261574 0.3220481 
```{r}
rfVarDrop <-rf_Imp_Sort
#trim the bottom x predictors
#rfVarDrop <- rfVarDrop %>% filter(rank <= 65)
rfVarDrop <- rfVarDrop %>% filter(rank <= 30)

dropVar <- row.names(rfVarDrop) #Variables to drop list 

#make a shorter list of variables from rf variable selection
dfTrim <- df0[ , ! names(df0) %in% dropVar] 
dfTrim <- na.omit(dfTrim)
#dfTrim is experimental list.

```

```{r}
#create new train and test data from rf selection
train2 <- sample(1:nrow(dfTrim), nrow(dfTrim)*0.7)
trim_train_data <- dfTrim[train2,]
trim_test_data <- dfTrim[-train2,]
#str(dfTrim)
```

```{r}
start_t <- Sys.time()
cat("",cat("Trimmed Training started at:",format(start_t, "%a %b %d %X %Y")))

rf_eval2 <- randomForest(hospital_death~., data = trim_train_data,
                           mtry = 5, importance = TRUE)

rf_eval2
varImpPlot(rf_eval2)

finish_t <- Sys.time()
cat("",cat("Trimmed Training finished at:",format(finish_t, "%a %b %d %X %Y")))

cat("Trimmed The training process finished in",difftime(finish_t,start_t,units="mins"), "minutes")
```

```{r} 
#Check accuracy score
rf_yhat2 <- predict(rf_eval2, newdata = trim_test_data)
#accuracy score of mtry 5 rf used for variable selection
rfscore2 <- postResample(rf_yhat2, trim_test_data$hospital_death)
rfscore2
#rfscore
```

 Accuracy     Kappa   lowest 50 predictors removed
0.9295177 0.3744659
 Accuracy     Kappa   lowest 60 predictors removed
0.9272813 0.3754597
Accuracy     Kappa    lowest 70 predictors removed
0.9203867 0.3451922
 Accuracy     Kappa    lowest 65 predictors removed
0.9316949 0.3691730 

```{r}
table(df$hospital_death)
```

```{r}
#Run Tune on RF Austin  took 95 minutes with 4:15
#tuneGrid <- data.frame(mtry = 4:15)
tuneGrid <- data.frame(mtry = 9:9)
#tuneGrid value came out 9.


control <- trainControl(method = 'cv', number = 5)
# print out system time before training
start_t <- Sys.time()
cat("",cat("RFtune Training started at:",format(start_t, "%a %b %d %X %Y")))

rftrim_tuned <- train(hospital_death ~ ., data = trim_train_data,
                  method = 'rf',
                  trControl = control,
                  tuneGrid = tuneGrid)

# print out system time after training
finish_t <- Sys.time()
cat("",cat("RFTune Training finished at:",format(finish_t, "%a %b %d %X %Y")))

cat(" RFTune The training process finished in",difftime(finish_t,start_t,units="mins"), "minutes")

print(rftrim_tuned)

#
```
RFtune Training started at: Fri Apr 29 5:13:10 PM 2022RFTune Training finished at: Fri Apr 29 6:48:32 PM 2022 RFTune The training process finished in 95.35552 minutesRandom Forest 

42996 samples
   50 predictor
    2 classes: '0', '1' 

No pre-processing
Resampling: Cross-Validated (5 fold) 
Summary of sample sizes: 34397, 34397, 34397, 34396, 34397 
Resampling results across tuning parameters:

  mtry  Accuracy   Kappa    
   4    0.9273421  0.3179540
   5    0.9275049  0.3288148
   6    0.9281328  0.3437036
   7    0.9281096  0.3470367
   8    0.9280398  0.3514676
   9    0.9283654  0.3562677
  10    0.9280165  0.3565548
  11    0.9282957  0.3599356
  12    0.9282259  0.3625530
  13    0.9280166  0.3602829
  14    0.9279003  0.3596171
  15    0.9282259  0.3656226

Accuracy was used to select the optimal model using the largest value.
The final value used for the model was mtry = 9.
```{r} 
#Check accuracy score of RF tuned
rf_yhat_tuned <- predict(rftrim_tuned, newdata = trim_test_data)
#accuracy score of mtry 5 rf used for variable selection
rfscore_tuned <- postResample(rf_yhat_tuned, trim_test_data$hospital_death)
rfscore_tuned

print(rfscore_tuned)
plot(rfscore_tuned)
```
 Accuracy     Kappa  65 removed
0.9317824 0.3733778  

```{r}
#rownames(rfVarDrop)
colnames(dfTrim)
```
Drop list: 
 [1] "bmi"                         "elective_surgery"            "ethnicity"                   "gender"                      "height"                     
 [6] "icu_admit_source"            "icu_id"                      "icu_stay_type"               "icu_type"                    "pre_icu_los_days"           
[11] "weight"                      "apache_2_diagnosis"          "apache_post_operative"       "arf_apache"                  "gcs_eyes_apache"            
[16] "gcs_motor_apache"            "gcs_unable_apache"           "gcs_verbal_apache"           "intubated_apache"            "map_apache"                 
[21] "resprate_apache"             "ventilated_apache"           "d1_diasbp_max"               "d1_diasbp_noninvasive_max"   "d1_mbp_max"                 
[26] "d1_mbp_min"                  "d1_mbp_noninvasive_max"      "d1_mbp_noninvasive_min"      "d1_resprate_max"             "d1_resprate_min"            
[31] "d1_spo2_max"                 "d1_sysbp_max"                "d1_sysbp_noninvasive_max"    "h1_diasbp_max"               "h1_diasbp_min"              
[36] "h1_diasbp_noninvasive_max"   "h1_diasbp_noninvasive_min"   "h1_heartrate_max"            "h1_heartrate_min"            "h1_mbp_max"                 
[41] "h1_mbp_min"                  "h1_mbp_noninvasive_max"      "h1_mbp_noninvasive_min"      "h1_resprate_max"             "h1_resprate_min"            
[46] "h1_spo2_max"                 "h1_spo2_min"                 "h1_sysbp_max"                "h1_sysbp_min"                "h1_sysbp_noninvasive_max"   
[51] "h1_sysbp_noninvasive_min"    "d1_glucose_max"              "d1_glucose_min"              "d1_potassium_max"            "d1_potassium_min"           
[56] "aids"                        "cirrhosis"                   "diabetes_mellitus"           "hepatic_failure"             "immunosuppression"          
[61] "leukemia"                    "lymphoma"                    "solid_tumor_with_metastasis" "apache_3j_bodysystem"        "apache_2_bodysystem"  

Keep List:
 [1] "age"                           "apache_3j_diagnosis"           "heart_rate_apache"             "temp_apache"                   "d1_diasbp_min"                
 [6] "d1_diasbp_noninvasive_min"     "d1_heartrate_max"              "d1_heartrate_min"              "d1_spo2_min"                   "d1_sysbp_min"                 
[11] "d1_sysbp_noninvasive_min"      "d1_temp_max"                   "d1_temp_min"                   "apache_4a_hospital_death_prob" "apache_4a_icu_death_prob"     
[16] "hospital_death"  
Tune parameters of favored model                                    	Rane
Validate favored model on Test group                                	Rane
```{r}
stopCluster(cl)
```
